AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference

### 2019-08-17 paper/learning/notes.tex
updated

\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{cancel}
\tikzset{>=latex}
\usetikzlibrary{bayesnet}
\usetikzlibrary{arrows,shapes,automata}
\usetikzlibrary{positioning}
\newcommand{\n}[1]{\ensuremath{n_{#1}}}
\newcommand{\rv}[1]{\ensuremath{{#1}_{1:n_{#1}}}}
\newcommand{\rvv}[1]{\ensuremath{{#1}_{1:n_{#1'}}}}
\newcommand{\model}[1]{\ensuremath{\mathcal{M}{(#1)}}}
\newcommand{\simu}[1]{\ensuremath{\Omega(\model{#1})}}
\newcommand{\micro}[2]{\ensuremath{\phi_{#1}(#2)}}
\newcommand{\pdf}[1]{\ensuremath{\mathcal{P}_{d}(#1)}}
\newcommand{\pmf}[1]{\ensuremath{\mathcal{P}_{m}(#1)}}
\newcommand{\add}[1]{\ensuremath{A_{#1}}}
\newcommand{\adds}[2]{\ensuremath{A_{#1}_{#2}}}
\newcommand{\expect}[1]{\ensuremath{\mathbb{E}[#1]}}
\newcommand{\trace}[1]{\ensuremath{\mathcal{T}_{#1}}}
\newcommand{\traces}[2]{\ensuremath{\textit{tr}^{#1}_{#2}}}
\newcommand{\stoc}[2]{\ensuremath{f^{#1}_{#2}(\rvv{x})}}
% \usepackage[binary-units=true]{siunitx}
\title{Hijacking Simulators with Universal Probabilistic Programming}

@@

The standard algorithm is defined as follows:
\begin{algorithm}
  \label{alg:rsstats}
  \small
   \caption{Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

@@

\noindent
\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
\hline
Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
\hline
\begin{lstlisting}[mathescape=true]
    def f():
      $z_{1} \sim \mathcal{N}(0,1)$
      $z_{2} \gets R_{1}(z_1, 1)$
      $z_{3} \sim \mathcal{U}(0,2)$
      $z_{4} \gets R_{2}(z_2,z_3, 1)$
    return $g(z_{1:4})$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_1(z_1)$:
  $x \sim \mathcal{N}(z_1,1)$
  if $x > 0$:
    return $x$
  else:
    return $R_{1}(z_1)$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_2(z_2, z_3)$:
  $y \gets 0$
  while $y < z_2$:
    $y \sim \mathcal{N}(z_3,1)$
  end
  return $y$
\end{lstlisting}\\
\hline
$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
\hline
\end{tabular}

@@

The framework we present in this section is generally applicable to stochastic simulator 
environment, where there exists rejection sampling loops.

We start by noting that the prior, defined in the context of the outer samples is 
defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
We start with the information projection we construct the variational objective as follows:
\begin{align*}
  J(\eta) = KL(p(x) || q(x; \eta)) \\
   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
  %  \int p(x')dx' \gets 1 \\
  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
\end{align*}

Thus, to define the objective for each subproblem $r  \in S_{R}$ as:
\begin{equation}
  \label{eq:objective}
  J_{r}(\eta_r) = -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j = r)\ln(q_r(x_j;\eta_r, \phi_j))]
\end{equation}

@@

\subsection{Method 3: Learn marginal likelihood $p(\cdot)$ approximation}

\begin{equation}
  p(x) = 
  \begin{cases}
  \prod^{n_x}_{j=1,{a_j}\in S_{R}} f_{a_j}(x_j ; \psi_j) \prod^{n_x}_{j=1,{a_j}\in S_{R}} \frac{P^{in}_{a_j}(x_j,\psi_j)}{P^{in}_{a_j}(\psi_j)} \\
  0
  \end{cases}
\end{equation}
The normalising constant inside (in) the rejection sampler is intractable.

>>>>

\usepackage{amsmath,amsthm}
\usepackage{amssymb}
\usepackage{cancel}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
\tikzset{>=latex}
\usetikzlibrary{bayesnet}
\usetikzlibrary{arrows,shapes,automata}
\usetikzlibrary{positioning}
\newcommand{\n}[1]{\ensuremath{n_{#1}}}
\newcommand{\rv}[1]{\ensuremath{{#1}_{1:n_{#1}}}}
\newcommand{\rvv}[1]{\ensuremath{{#1}_{1:n_{#1'}}}}
\newcommand{\model}[1]{\ensuremath{\mathcal{M}{(#1)}}}
\newcommand{\simu}[1]{\ensuremath{\Omega(\model{#1})}}
\newcommand{\micro}[2]{\ensuremath{\phi_{#1}(#2)}}
\newcommand{\pdf}[1]{\ensuremath{\mathcal{P}_{d}(#1)}}
\newcommand{\pmf}[1]{\ensuremath{\mathcal{P}_{m}(#1)}}
\newcommand{\add}[1]{\ensuremath{A_{#1}}}
\newcommand{\adds}[2]{\ensuremath{A_{#1}_{#2}}}
\newcommand{\expect}[1]{\ensuremath{\mathbb{E}[#1]}}
\newcommand{\trace}[1]{\ensuremath{\mathcal{T}_{#1}}}
\newcommand{\traces}[2]{\ensuremath{\textit{tr}^{#1}_{#2}}}
\newcommand{\stoc}[2]{\ensuremath{f^{#1}_{#2}(\rvv{x})}}
% \usepackage[binary-units=true]{siunitx}
\title{AIRS: Amortized Inference for Rejection Samplers for Fast Simulator Inference}

@@

The standard algorithm is defined as follows:
\begin{algorithm}
  \label{alg:rsstats}
  \small
   \caption{ABC: Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

but more generally speaking we can write a rejection sampler as follows:
\begin{algorithm}
  \label{alg:rsstats}
  \small
   \caption{Simulator: Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim p(x)$ 
       \State $\mathbb{I}(g(x) > 0)$
       \State $ Y \sim p(X=x | \mathbb{I}(g(x) > 0))$
       \State accept $X_i$ with $p(\mathbb{I}(g(x) > 0))$
\end{algorithmic}
\end{algorithm}
Thus, the probability of generating a sample $X = x$ from the true distribution can be defined, using Bayes rule, 
as: 
\begin{equation}
  p(X=x| \mathbb{I}) = \frac{p(\mathbb{I} | X=x)p(X=x)}{p(\mathbb{I})} 
\end{equation}
where $p(\mathbb{I} | X=x) = \{0,1\}$.

@@

\noindent
\begin{tabular}{|p{3.6cm}|p{3.6cm}|p{3.6cm}|}
\hline
Stochastic Simulator  & Rejection Sampler 1 & Rejection Sampler 2 \\
\hline
\begin{lstlisting}[mathescape=true]
    def f():
      $z_{1} \sim \mathcal{N}(0,1)$
      $z_{2} \gets R_{1}(z_1, 1)$
      $z_{3} \sim \mathcal{U}(0,2)$
      $z_{4} \gets R_{2}(z_2,z_3, 1)$
    return $g(z_{1:4})$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_1(z_1)$:
  $x \sim \mathcal{N}(z_1,1)$
  if $x > 0$:
    return $x$
  else:
    return $R_{1}(z_1)$
\end{lstlisting}&
\begin{lstlisting}[mathescape=true]
def $R_2(z_2, z_3)$:
  $y \gets 0$
  while $y < z_2$:
    $y \sim \mathcal{N}(z_3,1)$
  end
  return $y$
\end{lstlisting}\\
% \begin{lstlisting}[mathescape=true]
%   def $R_3(z_4)$:
%     $y \sim \mathcal{U}(0,1)$
%     $x \sim \mathcal{U}(0,1.5)$
%     while $y < z_2$:
%       $y \sim \mathcal{N}(z_3,1)$
%     end
%     return $y$
%   \end{lstlisting}\\
\hline
$P(\mathcal{T}= x_{1:n_x} | \lambda)$ & $q_{1}(z_{2} | z_{1})$ & $q_{2}(z_{4} |z_{2},z_{3})$ \\
\hline
\end{tabular}

@@

The framework we present in this section is generally applicable to any stochastic simulator 
environment, where there exists any form of rejection sampling blocks.
To perform the inferences that we require, we construct a variational approximation of the rejection sampling 
blocks via an amortization procedure
and note that the variational approximation in the context of our setting is specified by three things:
\begin{enumerate}
  \itemsep0em 
  \item Makes use of existing parts of the outer program including all deterministic operations.
  \item Form of the $\gamma_{a_j}(x_j ; \eta_{a_j}, \psi_j) \forall a_j \in S_r$. 
  \item There exists variational parameters $\eta_{a_j} \forall a_j \in S_r$.
\end{enumerate}
where $a_j \in \{1, \ldots, L\}$ corresponds to the set of all addresses, $S_r$ denotes the set of addresses $a_j$ that correspond directly to the locations 
of the rejection sampling blocks, that is $a_j \cup S_r = \{1, \ldots, L\}$, $\gamma_{a_j}$ denotes the variational approximation at address $a_j$, 
for example, $\gamma_{a_j} = \mathcal{N}(x_j ; \mu_{a_j}(\psi_j; \eta_{a_j}), \Sigma_{a_j}(\psi_j ; \eta_{a_j}))$, with $\eta_{a_j}$ representing the variational
parameters and $\psi_j$ representing all variables that input into the given program.

We start by noting that the prior, defined in the context of the outer samples is defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
next, we state for a given program $\mathcal{P}$ we denote the proposal for the program as:
\begin{equation}
  q(\mathbf{x}; \kappa) = 
  \begin{cases}
  \prod^{n_x}_{j=1, j\notin S_{r}} f_{a_j}(x_j ; \psi_j)(\prod^{n_x}_{j=1, j \in S_{r}} \gamma_{a_j}(x_j; \eta_{a_j}, \psi_{j})) \texttt{ if } \mathcal{B}(x_{1:n_{x}} = 1)\\
  0
  \end{cases}
\end{equation}
where $\kappa = \{\eta_{a_j} ; a_j \in S_{r}\}$ are the variational parameters.
As we now have the density of the proposal and the density of the program we can construct a variational objective as follows:

using the information projection we construct the variational objective as follows:
\begin{align*}
  J(\eta) = KL(p(x) || q(x; \kappa)) \\
   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
  %  \int p(x')dx' \gets 1 \\
  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
   \kappa^{*} = \operatorname*{argmin}_{\kappa} \mathbb{E}_{X \sim p(x)}[-\log(q(X=x; \kappa))] \\
   = \operatorname*{argmax}_{\kappa} \mathbb{E}_{X \sim p(x)}[\sum^{n_x}_{j=1, j \in S_r} \log( \gamma_{a_j}(X_j = x_j ; \eta_{a_j}, \psi_j))] \\
   \forall r \in S_r, \eta^{*}_r = \operatorname*{argmax}_{r} \mathbb{E}_{x \sim p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j =r) \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j))] \\
   \nabla_{\kappa} J(\kappa) = \nabla_{\kappa} \mathbb{E}_{X \sim p(x)}[\sum^{n_x}_{j=1}\mathbb{I}(a_j =r) \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j))] \\
   = \mathbb{E}_{X \sim p(x)}[ \sum^{n_x}_{j=1, j \in S_r} \nabla_\kappa \log(\gamma_r (X_j = x_j ; \eta_r, \psi_j)]\\
   \approx \frac{1}{N} \sum^{N}_{n=1} \sum^{n_x}_{j=1, j\in S_r} \nabla_{\kappa}\log(\gamma_r (x^{n}_j ; \eta_r, \psi_j))
  %  = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
\end{align*}

Thus, to define the objective for each subproblem $r  \in S_{r}$ as:
\begin{equation}
  \label{eq:objective}
  \nabla_{\eta_r} J(\kappa) \approx \frac{1}{N} \sum^{N}_{n=1} \sum^{n_x}_{j=1} \mathbb{I}(r=a_j) \nabla_{\eta_r}\log(\gamma_r (x^{n}_j ; \eta_r, \psi_j))
\end{equation}

where $x^{n}_{1:n_x} \stackrel{iid}\sim p(x)$.

@@

\subsection{Method 3: Learn marginal likelihood approximation}
With $ p(\cdot)$ representing the marginal likelihood. 
\begin{equation}
  p(x) = 
  \begin{cases}
  \prod^{n_x}_{j=1,{a_j}\in S_{R}} f_{a_j}(x_j ; \psi_j) \prod^{n_x}_{j=1,{a_j}\in S_{R}} \frac{P^{in}_{a_j}(x_j,\psi_j)}{P^{in}_{a_j}(\psi_j)} \\
  0
  \end{cases}
\end{equation}
The normalising constant inside (in) the rejection sampler is intractable.
